\chapter{Discrete model}
\label{ch:discrete-model}

\epigraph{... all models are approximations. Essentially, all models are wrong, but some
are useful. However, the approximate nature of the model must always be borne in
mind....}{Empirical Model-Building and Response Surfaces, 1987---George E. P.  
Box}

The mathematical model is discretized in algebraic operations, in order to be 
computable.

\section{Charge assignment}
At each grid point $g$ at $\x$ we accumulate the charge of each particle $p$ in 
$\x_p$ as
%
\begin{equation}%{{{
\rho(\x) = \sum_p q\,W(\x - \x_p) + \rho_0
\end{equation}%}}}
%
The background charge density $\rho_0$ is used to neutralize the total charge 
when is non-zero. The weighting function $W$ determines the shape of the 
particle charge. Different schemes can be used to approximate the charge density 
from the particles. We will focus on bilinear interpolation for it's simplicity 
and low computation requirements. The corresponding weighting function can be 
written as
%
\begin{equation}%{{{
W(\x) =
\begin{cases}
			\displaystyle\left(1 - \frac{|x|}{\Delta x}\right)
				\left(1 - \frac{|y|}{\Delta y}\right) & \text{if}\ -\Delta\x < \x < 
				\Delta \x\\
			0 & \text{otherwise}
\end{cases}
\end{equation}%}}}
%
Notice that a particle $p$ always affects the four enclosing grid points in the 
neighbourhood $\neigh{p}$, but more complex interpolation methods may extend the 
update region even further. It may be noted that the increase in smoothing, at 
computation expense, can gain from the reduced number of particles needed to 
obtain a similar result, avoiding nonphysical effects.
%
\begin{figure}[]%{{{
\centering
\begin{tikzpicture}[
		>=latex,
		effect/.style={dashed,-{Latex[length=3mm, width=1mm]}},
		particle/.style={fill=black,radius=3pt},
	]
	\draw [step=2cm,dotted] (1,1) grid (5,5);
	\coordinate (p) at (2.5,3.2);
	\coordinate (center) at (3,3);
	\coordinate (A) at ($(center)+(-1,1)$);
	\coordinate (B) at ($(center)+(1,1)$);
	\coordinate (C) at ($(center)+(1,-1)$);
	\coordinate (D) at ($(center)+(-1,-1)$);
	\draw[effect] (p) -- (A);
	\draw[effect] (p) -- (B);
	\draw[effect] (p) -- (C);
	\draw[effect] (p) -- (D);
	\node[above left]  at (A) {$A$};
	\node[above right] at (B) {$B$};
	\node[below right] at (C) {$C$};
	\node[below left]  at (D) {$D$};
	\draw[particle] (p) circle;
	\node[left] at (p) {$p$};
\end{tikzpicture}
\hspace{0.5cm}
\begin{tikzpicture}[
		>=latex,
		box/.style={black},
		particle/.style={fill=black,radius=3pt},
		div/.style={dashed},
	]
	\draw [step=2cm,dotted] (1,1) grid (5,5);
	\coordinate (p) at (2.5,3.2);
	\coordinate (center) at (3,3);
	\coordinate (A) at ($(p)+(-1,1)$);
	\coordinate (B) at ($(p)+(1,1)$);
	\coordinate (C) at ($(p)+(1,-1)$);
	\coordinate (D) at ($(p)+(-1,-1)$);
	\draw[box] (A) -- (B) -- (C) -- (D) -- (A);
	\draw[div] ($(A)!(center)!(B)$) -- (center);
	\draw[div] ($(B)!(center)!(C)$) -- (center);
	\draw[div] ($(C)!(center)!(D)$) -- (center);
	\draw[div] ($(D)!(center)!(A)$) -- (center);
	\node at ($(center)!0.5!(A)$) {$a$};
	\node at ($(center)!0.5!(B)$) {$b$};
	\node at ($(center)!0.5!(C)$) {$c$};
	\node at ($(center)!0.5!(D)$) {$d$};
	\draw[particle] (p) circle;
\end{tikzpicture}
\hspace{0.5cm}
\begin{tikzpicture}[
		>=latex,
		box/.style={black},
		particle/.style={fill=black,radius=3pt},
		div/.style={dashed},
	]
	\draw [step=2cm,dotted] (1,1) grid (5,5);
	\coordinate (p) at (2.5,3.2);
	\coordinate (center) at (3,3);
	\coordinate (A) at ($(center)+(-1,1)$);
	\coordinate (B) at ($(center)+(1,1)$);
	\coordinate (C) at ($(center)+(1,-1)$);
	\coordinate (D) at ($(center)+(-1,-1)$);
	\draw[box] (A) -- (B) -- (C) -- (D) -- (A);
	\draw[div] ($(A)!(p)!(B)$) -- (p);
	\draw[div] ($(B)!(p)!(C)$) -- (p);
	\draw[div] ($(C)!(p)!(D)$) -- (p);
	\draw[div] ($(D)!(p)!(A)$) -- (p);
	\node at ($(p)!0.5!(A)$) {$c$};
	\node at ($(p)!0.5!(B)$) {$d$};
	\node at ($(p)!0.5!(C)$) {$a$};
	\node at ($(p)!0.5!(D)$) {$b$};
	\draw[particle] (p) circle;
\end{tikzpicture}
\caption{Interpolation of particle $p$ charge into the four grid points A to D.}
\label{fig:interpolation}
\end{figure}%}}}
%
The particle $p$ has a uniform charge area, centered at the particle position 
$\x_p$, with size $\Delta \x$, as shown in the figure~\ref{fig:interpolation}.  
Each grid point $A,B,C$ and $D$ receives the amount of charge weighed by the 
area $a,b,c$ and $d$. It can be observed that the area is equal to the opposite 
region, when the particle $p$ is used to divide the grid cell.
%
The particle shape can be altered later in the Fourier space, without large 
computation effort, in case the solver already computes the FFT.
%
\section{Field equations}

In order to compute the electric field $\E$, the electric potential $\phi$ is 
generally needed, which can be obtained from the charge density $\rho$.

\subsection{Electric potential}
Several methods are available to solve the Poisson equation 
(Eq.~\ref{eq:poisson}).

\paragraph{Iterative  methods} such as Jacobi, Gauss-Seidel, Successive Over 
Relaxation (SOR), Chebyshev acceleration are some of the most familiar methods 
to solve the Poisson equation.

\paragraph{Matrix methods} The equations from finite differencing the mesh are 
considered a large system of equations. We can find in this methods the Thomas 
Tridiagonal algorithm, Conjugate-Gradient, LU or Incomplete Decomposition.

\paragraph{Spectral methods} Also known as Rapid Elliptic Solvers (RES) are a 
family of methods that use the fast Fourier transform (FFT). Are know for being 
usually faster than the previous ones, with a complexity in $O(N_g \log_2 N_g)$

\vspace{1em}
\noindent
%
We will only focus on the LU for small problems and for testing, and spectral 
methods, more specific on the Multiple Fourier Transform (MTF) method, as it is 
the main method implemented in the simulator, due to its relative simplicity and 
low computational complexity.

\subsection{LU decomposition}
%
% TODO: Check the error bound
For two dimensions, we can approximate the solution using the second order 
centered finite differences (with an error proportional to $\Delta x ^2 \Delta 
y^2$), as
%
\begin{equation}%{{{
\label{eq:discrete-poisson}
\frac{\phi(x-1, y) + \phi(x, y-1) - 4\phi(x,y) + \phi(x+1,y)+\phi(x,y+1)}{\Delta 
x ^2 \Delta y^2} = - \frac{\rho(x,y)}{\epsilon_0}
\end{equation}%}}}
%
which leads to a system of $N_g$ linear equations and can be also written in 
matrix form
%
\begin{equation}
\label{eq:eq-system}
A\phi = -\frac{\Delta x ^2 \Delta y^2\,\rho}{\epsilon_0}
\end{equation}
%
The $N_g \times N_g$ coefficient matrix $A$ has non-zero coefficients only at 
$a_{ii} = 4$ and $a_{ij} = -1$ with $j \in \{i+1, i-1, i+N_x, i-N_x\} \mod N_x$, 
for all $0 \le i \le Ng$.
%
However, the matrix $A$ is singular, so the system of equations has infinite 
solutions. Boundary conditions can be added to get a unique solution. The extra 
equation $\phi(0,0) = 0$ leads to a system with only one solution, but with one 
extra equation. In order to keep the matrix $A$ square, the following steps may 
be taken:

\begin{enumerate}
\item Subtract  the extra equation $\phi(0,0) = 0$ to the first row of $A$, with 
the only change in the coefficient to $a_{11} = 3$.

\item Add all first $N_g$ equations: Each equation has one coefficient of $4$ 
and four of $-1$ except the first equation. Also we assume the total charge 
density is zero, obtaining $\phi(0,0) = 0$.

\item Subtract it from the last equation, which leads to a zero coefficient that 
can be removed.
\end{enumerate}
%
The only change that remains is at the coefficient $a_{11} = 3$. Now the matrix 
$A$ is squared and non-singular and has only one solution and can now be solved 
with the $LU$ method.

The $LU$ decomposition, with a complexity in $O(2/3N_g^3)$, can be used to form 
two systems of equations that can be solved faster. If we rewrite the system of 
equations~\ref{eq:eq-system} as the usual form $Ax=b$ with
\begin{equation}
x = \phi,\quad b = -\frac{\Delta x ^2 \Delta y^2\,\rho}{\epsilon_0}
\end{equation}
%
Then we can use the decomposition $A=LU$ to form two systems of equations
%
\begin{equation}
\label{eq:LU-systems}
Ux=y, \quad Ly=b
\end{equation}
%
which can be solved in complexity $O(2N_g^2)$.


\subsection{Multiple Fourier Transform (MFT)}

The general second-order PDE with constant coefficients and periodic boundary 
conditions
%
\begin{equation}%{{{
\label{eq:gen-fd}
a \frac{\partial^2 \phi}{\partial x^2}+b\frac{\partial \phi}{\partial x}+c\phi +
d \frac{\partial^2 \phi}{\partial y^2}+e\frac{\partial \phi}{\partial y}+f\phi = 
g(x,y)
\end{equation}%}}}
%
can be solved by using the FFT. If we expand $\phi$ and $g$ in a finite double 
Fourier series, we obtain
%
\begin{equation}%{{{
\phi(x,y) = \sum_{k,l} \hat \phi(k, l) \exp\left({\frac{2\pi i (xk + 
yl)}{n}}\right)
\end{equation}%}}}
%
and
%
\begin{equation}%{{{
g(x,y) = \sum_{k,l} \hat g(k, l) \exp\left({\frac{2\pi i (xk + yl)}{n}}\right)
\end{equation}%}}}
%
which now can be substituted in the Eq.~\ref{eq:gen-fd}, to obtain
%
\begin{equation}%{{{
\hat \phi(k,l) = \hat G(k,l) \, \hat g(k,l),\quad 0<k<N_x,\,0<l<N_y
\end{equation}%}}}
%
with for a unit mesh
%
\begin{equation}%{{{
\begin{split}
\hat G(k,l) = \Bigg[
& 2a \left( \cos \frac{2\pi k}{n} - 1 \right) +
ib \sin \frac{2\pi k}{n} + c \,+ \\
& 2d \left( \cos \frac{2\pi l}{n} - 1 \right) +
ie \sin \frac{2\pi l}{n} + f
\Bigg]^{-1}
\end{split}
\end{equation}%}}}
%
To solve the Poisson equation, discretized as Eq.~\ref{eq:discrete-poisson}, we 
have $a=d=1$ and $b=c=e=f=0$ so we can simplify $\hat G(k,l)$ as
%
\begin{equation}%{{{
\hat G(k,l) = \frac{1}{2}\left[
\cos \frac{2\pi k}{n} +
\cos \frac{2\pi l}{n} -
2 \right]^{-1}
\end{equation}%}}}
%
Let $g = -{\Delta x ^2 \Delta y^2\,\rho}/{\epsilon_0}$, then the steps to 
compute the electric potential can be summarized as follows:
%
\begin{center}%{{{
\begin{tikzpicture}[>=latex,thick]
	\matrix (m) [
		matrix of nodes,
		column sep=20mm,
		nodes={
			%line width=1pt,
			anchor=center,
			text centered,
			%minimum width=1cm,
			minimum height=8mm,
		},
%		txt/.style={text width=1.5cm,anchor=center},
	]
	{
		$g$ & $\hat g$ & $\hat \phi$ & $\phi$ \\%& $\E$\\
	};
	\foreach \i [evaluate={\j=int(\i+1)}] in {1,...,3}{
		\draw[->] (m-1-\i) -- (m-1-\j);
	}
	\draw[draw=none] (m-1-1) -- (m-1-2) node[midway,above] {FFT};
	\draw[draw=none] (m-1-2) -- (m-1-3) node[midway,above] {$\hat G$};
	\draw[draw=none] (m-1-3) -- (m-1-4) node[midway,above] {IFFT};
%	\draw[draw=none] (m-1-4) -- (m-1-5) node[midway,above]
%	{Eq.~\ref{eq:phi-to-E}};
\end{tikzpicture}
\end{center}%}}}
%
\begin{enumerate}
\item Compute the complex FFT $\hat g$ of $g$
\item Multiply each element of $\hat g$ by the corresponding complex coefficient 
$\hat G$, to obtain $\hat \phi$
\item Compute the inverse FFT of $\hat \phi$ to get $\phi$
\end{enumerate}
%
The complexity in the worst case is in $O(N_g \log_2 N_g)$ with the number of 
total points in the grid $N_g$.

\subsection{Electric field}
The electric field $\E$ can then be obtained by centered first order finite 
differences in each dimension
%
\begin{equation}%{{{
\begin{split}
\label{eq:phi-to-E}
\E_x(x,y) &= \frac{\phi(x-1,y) - \phi(x+1,y)}{2\,\Delta x} \\
\E_y(x,y) &= \frac{\phi(x,y-1) - \phi(x,y+1)}{2\,\Delta y}
\end{split}
\end{equation}%}}}
%

\section{Force interpolation}

The force acting on a particle $p$ can be decomposed in two main parts, the 
electric and magnetic force $\F=\F_E + \F_B$.

The electric force $\F_E$ is computed similarly as the charge deposition, but in 
the reverse order. The force $\F_E$ is interpolated from the electric field $\E$ 
of the neighbour grid points $\neigh{p}$, using the same interpolation function 
$W$.
\begin{equation}%{{{
\begin{split}
\F_E &= q \sum_{g \in \neigh{p}} W(\x_p - \x_g)\ \E(\x_g) \\
\end{split}
\end{equation}%}}}
Notice that a particle $p$ only needs the values of the electric field in the 
neighbourhood $\neigh{p}$.

The magnetic force $\F_B$ is constant in the simulator, as we only consider a 
fixed background magnetic field $\B_0$. For a particle $p$ with velocity $\v$  
can be written as
%
\begin{equation}%{{{
\F_B = q (\v \times \B_0)
\end{equation}%}}}


\section{Equations of motion}

In order to move the particles, the equations of motion need to be solved:
%
\begin{equation}
\frac{d\x}{dt}=\v
\end{equation}
\begin{equation}
m \frac{d\v}{dt} = \F
\end{equation}
%
The \textit{leap-frog} method is a common integration scheme with second-order 
accuracy and an error proportional to $\Delta t^2$. The name describes de 
behavior of the position and velocity, which are updated at interleaved time 
steps, similarly to the trajectory of a frog. The method is time reversible with 
an stability far superior of other higher-order integration methods, such as 
fourth order Runge-Kutta. A more in depth stability analysis can be found in 
Chapter 4 of Hockney and Eastwood book~\cite{hockney}.  The discretized 
equations can be written as
%
\begin{equation}
\frac{\x^{n+1} - \x^{n}}{\Delta \x} = \v^{n + 1/2}
\end{equation}
%
\begin{equation}
m\frac{\v^{n+1/2} - \v^{n-1/2}}{\Delta \x} = \F(x^n)
\end{equation}
%
Several methods are available, but we will focus on the Boris integrator.

\subsection{Boris integrator}

Consists of three steps:
%
\begin{enumerate}
\item Add half of the electric impulse
\item Rotate
\item Add the remaining half electric impulse
\end{enumerate}
%
The Boris integrator computes the velocity of a particle in a constant electric 
field $\E$ and a constant magnetic field $\B$. We have the velocity 
$\v_{t-\Delta t/2}$ of the particle at $t-\Delta t/2$ as we use the leapfrog 
integrator.
%
\begin{figure}[h]
\centering
\begin{tikzpicture}[
	scale=2,
	>=latex]

	\def\centerarc[#1](#2)(#3:#4:#5)% Syntax: [draw options] (center) (initial angle:final angle:radius)
		{\draw[#1] ($(#2)+({#5*cos(#3)},{#5*sin(#3)})$) arc (#3:#4:#5); }

	\def\startangle{-25}
	\def\midangle{0}
	\def\endangle{25}
	\def\radius{2.0}
	\pgfmathsetmacro{\vlen}{\radius*tan(\startangle)}%

	\coordinate (O) at (0,0);
	\coordinate (S) at (\startangle:\radius);
	\coordinate (E) at (\endangle:\radius);

%	\centerarc[dashed](O)(\startangle:\endangle:\radius);
	\centerarc[->](O)(\startangle:\endangle:0.2*\radius);

	\draw (O)+(0.4,0.1) node [right] {$\theta$};

	\draw [thick,->] (O) -- (E) node [midway, above] {$\V{v^+}$};
	\draw [thick,->] (O) -- (S) node [midway, below] {$\V{v^-}$};

	\path (S) +(\startangle-90:\vlen) coordinate (V1E);
%	\path (E) +(\endangle-90:\vlen) coordinate (V3E);

%	%\draw [->] (E) -- (V3E);
	\draw [->] (S) -- (V1E) node [midway, right] {$\V{v'} \times \V t$};
%
	\draw [->] (O) -- (V1E) node [midway, above] {$\V{v'}$};
	\draw [->] (S) -- (E) node [near end, right] {$\V{v'} \times \V s$};

%	\draw [fill=white] (O) circle (0.02);

\end{tikzpicture}
\caption{Velocity space rotation from $\v-$ to $\v+$}
\end{figure}
%
\paragraph{Add half electric impulse} We define $\V{v^-}$ as the velocity after 
half a electric impulse:
$$\v^- = \v_{t-\dt/2} + \frac{q \E}{m} \frac{\dt}{2}$$

\paragraph{Rotate for the magnetic field} The rotation is done in two steps, 
first the half rotation is computed, with an angle of $\theta/2$:
$$\v' = \v^- + \v^- \times \V t $$

Then the rotation is completed by symmetry, using the $\V s$ vector
$$ \V s = \frac{2 \V t}{1 + \V t^2} $$
as
$$ \V{v^+} = \V{v^-} + \V{v}' \times \V{s} $$
